Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-07-06 manuscript/reviews/ResponseReview1.Rmd

> 8. Table 2: It could be helpful to show an example of a trial where an event never
occurs, to help the reader connect what is written in the text (line 325-327) to
what appears in the table.

Response: We added data visualizations in section A of the Supplemental Material.

> 9. Figure 4: Why is there no bin 40?

Response: Because no responses occurred in the first bin with endpoint 40 ms. However, we added this bin in Figure 2 (previously 4) of the revised ms.

@@

Response: We discuss the main contributions above and beyond existing works on page 6 of the revised ms. We address the issue of treating time as discrete or continuous on pages 8-9 of the revised ms. We removed redundancies, get to the tutorial section quicker (page 12) and included more precise explanations of technical jargon.

> <b>Code for the tutorials:</b> I went to the github page and found a folder for each of the 4 tutorials, which included sub-folders for data, figures, and tables. It took a minute for me to realize that the actual R code for each tutorial was in the main folder, not within each sub-folder. This may also be confusing to readers. Additionally, each of the 4 tutorials, along with their sub-tutorials are provided in separate .Rmd files. There were several download steps needed to get all the information. To me, this made the user experience tedious, and made the tutorials seem disjointed from one another. The authors might consider a more comprehensive single Rmd file that is neatly organized with an interactive table of contents (in the knitted html file). To me, this would make the experience much more enjoyable and useful for readers. Given that this is a tutorial, it would also be nice if the the RMarkdowns were somewhat more standalone e.g., by including a brief description of what a life table is and how to interpret the plot.

Response: We explain the location of the files and the folder structure in the README file. In addition we make it clear that a single download step is possible by either cloning the project, or downloading the ZIP file.
We also included a brief description of a life table and the definitions of the discrete-time functions in the R Markdown file called Tutorial_1a.Rmd.
At this point, we also prefer a modular approach to the tutorials rather than combining them. 
We appreciate that not everyone might share this preference, but we think there are many benefits to a modular workflow in this instance.

@@

> The numbered list on pg. 11 contained jargon statements without explanations. Suggest to simplify and keep the focus on the data you are using and specific research questions you are using survival analysis as a tool to address.

Response: We removed the numbered list in the revised ms.

> In section 2.3, suggest to lead with the data and research questions of interest and then show how features of survival analysis make it a good method to examine the research questions. I also felt that up until now, the tutorial has been very confusing and hard to follow. It could be that starting the tutorial at section 2.3 (with a few extra details filled in), may be an alternative.

Response: Section 2.3 has been rewritten. The new Figure 1 in the tutorial directly focuses  on the comparison between analysing means vs. distributional shapes. We also discuss general research questions on page 10 of the revised ms.

@@

> It wasn’t clear how section 3.1 added value to the tutorial

Response: Section 3 has been removed.

> Section 3.2 – it would be helpful if there was clarification in the text on when it would not be possible to include varying intercepts and slopes (i.e., at least clarifying that random intercepts are a requirement, but sometimes random slopes are removed based on model convergence issues – however, this also begs the question of whether a model should even be presented if it does not converge when it includes random slopes…)

Response: We do not discuss frequentist approaches in the main text anymore (to save space) and only mention that the reader can consult Tutorials 3a and 3b. We agree that a model should not be presented if it does not converge when it includes random slopes. However, to save space, and because this is an active area of reserch, we do not attempt to clarify when it would not be possible to include varying intercepts and/or slopes.

@@

> pg. 22 discusses warning messages and lets readers know these can be ignored but does not give details as to why. Since this is a didactic tutorial, it may be helpful to have at least some explanation as to why the warning can be ignored.

Response: On line 362 it was stated that "the warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored".
We now removed these messages from the paper and mention them now only in the tutorial itself.

> pg. 22 – since this is a tutorial, it might also be helpful to unpack in more detail the section about identifying individuals who may be guessing etc – especially since these are very important considerations.

Response: We have no more detail to share on this issue. On page 19 of the revised ms. we now write "In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may not be following task instructions (e.g., a flat conditional accuracy function at .5 indicates that someone is just guessing), outlying individuals, and/or different groups with qualitatively different behavior.".

### 2025-07-06 manuscript/reviews/ResponseReview1.tex

Response: We discuss the main contributions above and beyond existing
works on page 6 of the revised ms. We address the issue of treating time
as discrete or continuous on pages 8-9 of the revised ms. We removed
redundancies, get to the tutorial section quicker (page 12) and included
more precise explanations of technical jargon.

\begin{quote}
Code for the tutorials: I went to the github page and found a folder for
each of the 4 tutorials, which included sub-folders for data, figures,
and tables. It took a minute for me to realize that the actual R code
for each tutorial was in the main folder, not within each sub-folder.
This may also be confusing to readers. Additionally, each of the 4
tutorials, along with their sub-tutorials are provided in separate .Rmd
files. There were several download steps needed to get all the
information. To me, this made the user experience tedious, and made the
tutorials seem disjointed from one another. The authors might consider a
more comprehensive single Rmd file that is neatly organized with an
interactive table of contents (in the knitted html file). To me, this
would make the experience much more enjoyable and useful for readers.
Given that this is a tutorial, it would also be nice if the the
RMarkdowns were somewhat more standalone e.g., by including a brief
description of what a life table is and how to interpret the plot.
\end{quote}

Response: We explain the location of the files and the folder structure
in the README file. In addition we make it clear that a single download
step is possible by either cloning the project, or downloading the ZIP
file. We also included a brief description of a life table and the
definitions of the discrete-time functions in the R Markdown file called
Tutorial\_1a.Rmd. At this point, we also prefer a modular approach to
the tutorials rather than combining them. We appreciate that not
everyone might share this preference, but we think there are many
benefits to a modular workflow in this instance.

@@

\begin{quote}
The numbered list on pg. 11 contained jargon statements without
explanations. Suggest to simplify and keep the focus on the data you are
using and specific research questions you are using survival analysis as
a tool to address.
\end{quote}

Response: We removed the numbered list in the revised ms.

\begin{quote}
In section 2.3, suggest to lead with the data and research questions of
interest and then show how features of survival analysis make it a good
method to examine the research questions. I also felt that up until now,
the tutorial has been very confusing and hard to follow. It could be
that starting the tutorial at section 2.3 (with a few extra details
filled in), may be an alternative.
\end{quote}

Response: Section 2.3 has been rewritten. The new Figure 1 in the
tutorial directly focuses on the comparison between analysing means
vs.~distributional shapes. We also discuss general research questions on
page 10 of the revised ms.

@@

\begin{quote}
pg. 22 discusses warning messages and lets readers know these can be
ignored but does not give details as to why. Since this is a didactic
tutorial, it may be helpful to have at least some explanation as to why
the warning can be ignored.
\end{quote}

Response: On line 362 it was stated that ``the warning messages are
generated because some bins have no hazard and ca(t) estimates, and no
error bars. They can thus safely be ignored''. We now removed these
messages from the paper and mention them now only in the tutorial
itself.

\begin{quote}
pg. 22 -- since this is a tutorial, it might also be helpful to unpack
in more detail the section about identifying individuals who may be
guessing etc -- especially since these are very important
considerations.
\end{quote}

Response: We have no more detail to share on this issue. On page 19 of
the revised ms. we now write ``In general, it is important to visually
inspect the functions first for each participant, in order to identify
individuals that may not be following task instructions (e.g., a flat
conditional accuracy function at .5 indicates that someone is just
guessing), outlying individuals, and/or different groups with
qualitatively different behavior.''.

### 2025-07-06 manuscript/manuscript.Rmd
recreate pdfs after updating R

# R references
 r_refs(file = "r-references.bib", append=F) # append=F prevents multiple copies of the same citation/ creates a .bib-reference file
 my_r_citation <- cite_r(file = "r-references.bib", footnote = FALSE)
```

>>>>

# R references
 r_refs(file = "r-references.bib") #, append=F) # append=F prevents multiple copies of the same citation/ creates a .bib-reference file
 my_r_citation <- cite_r(file = "r-references.bib", footnote = FALSE)
```

### 2025-07-06 manuscript/manuscript.tex
recreate pdfs after updating R

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.4.0; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.1; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.13.1; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021), \emph{rstan} (Version 2.32.6; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.4; Barth, 2023) for all reported analyses.

>>>>

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.5.1; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.13.0; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.9.0.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.2; Wickham, 2016), \emph{lme4} (Version 1.1.37; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.4; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.3; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.168; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.4; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.14; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021), \emph{rstan} (Version 2.32.7; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.3.0; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.5; Barth, 2023) for all reported analyses.
